---
title: '10X Biological interpretation - Mouse Dataset'
author: "James Kinchen, University of Oxford"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
        code_folding: hide
        dev: [png,pdf]
params:
  script_name:
    input: text
    label: 'Script identifier'
    value: 'cluster_biology'
  data1:
    input: text
    label: 'Seurat dataset 1 name'
    value: '10X_Ms_Healthy'
  data2:
    input: text
    label: 'Seurat dataset 2 name'
    value: '10X_Ms_DSS'
---

```{r setup, include=FALSE, cache=FALSE}

knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 6, message = FALSE, warning = FALSE, cache = TRUE, 
                      cache.path = paste0("cache/",params$script_name,"/"),
                      fig.path = paste0("output/",params$script_name,"/"))

```

```{r setup_2, include=FALSE, cache=FALSE}

options(stringsAsFactors = FALSE)
library(Seurat)
library(dplyr)
library(Matrix)
library(colorspace)
library(viridis)
library(stats)
library(cowplot)
library(reshape2)
library(WGCNA)
library(gplots)
library(scater)

source("seurat_functions.R")

```

# Healthy dataset

## Data import and setup

```{r load_healthy}

# Load the preprocessed seurat object
sc_hc <- readRDS(paste0("output/clust_analysis/", params$data1, "/sc_reclustered.RData"))
clan_hc <- readRDS(paste0("output/clust_analysis/", params$data1, ".RData"))

```

## Identify and merge low confidence nodes

```{r node_confidence, fig.width=6, fig.height=4}

PlotClusterTree(sc_hc)

ns_df <- clan_hc$node.scores
ns_df$node <- factor(ns_df$node, levels = ns_df$node)
ggplot(data=ns_df, aes(x=node, y=oobe)) + geom_col() + geom_hline(yintercept = 0.05, linetype = "dashed")

```

```{r TSNE_split, fig.width=6, fig.height=4}

ColorTSNESplit(sc_hc, node = 28)
ColorTSNESplit(sc_hc, node = 24)

```

```{r merge_low_conf_hc}

sc_hc <- MergeNode(sc_hc, node.use = 24)
sc_hc <- BuildClusterTree(sc_hc, do.reorder = T, reorder.numeric = T, do.plot = F)
PlotClusterTree(sc_hc)
TSNEPlot(sc_hc, do.label = T, pt.size = 0.6)

```

## Identify clusters based on known marker expression

```{r marker_dotplot, fig.width=10, fig.height=8}

marker_list <- c("Epcam","Krt19","Kit","Ano1","Myh11","Des",
                 "Rgs5","Pdgfrb","Pecam1","Lyve1",
                 "S100b","Gfap","Cd52","Ptprc",
                 "Acta2","Dpt","Col6a2","Col1a2")

marker_spec <- c("Epithelial","Epithelial","ICC","ICC","Smooth muscle",
                 "Smooth muscle","Pericyte","Pericyte","Endothelial","Lymphatic",
                 "Glial","Glial","Immune","Immune","Fibroblast",
                 "Fibroblast","Fibroblast","Fibroblast")

cluster_labels <- c("Epithelial","Smooth muscle / ICC","Pericyte",
                    "Myofibroblast","SMA+ fibroblast", 
                    "Endothelium","Lymphatic","Glial","Immune",
                    "\n\n\nFibroblast subtypes\n\n\n")

dp_data <- calc_dots(sc_hc, marker_list)
dp_data <- dp_data[dp_data$cex > 0.05,]

p1 <- ggplot(data = dp_data) + 
  geom_point(aes(x=x,y=y,fill=col,size=cex), pch = 21) + 
  scale_fill_viridis(option = "inferno", "avg\nexp") + 
  scale_size_continuous(limits = c(0.05,1), "prob\ndetect") +  
  ylab("Cluster ID") +
  annotate("label", x = rep(23,10), y = c(1:9,11.5), label = cluster_labels, hjust = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90,hjust = 1, vjust = 0.5,size = 12), 
        axis.text.y = element_text(size=12),
        axis.title.x = element_blank(),
        legend.position = "left")

marker_spec <- factor(marker_spec, levels = unique(marker_spec))

p2 <- ggplot(data=data.frame(x=(1:18)-0.5, ms = marker_spec)) + geom_tile(aes(x=x,y=1,fill=ms)) + 
  scale_fill_discrete("Marker\nspecificity") +
  coord_cartesian(xlim = c(1,21.5)) +
  theme_nothing() + theme(legend.position = "left", plot.margin = unit(c(0,0,2,0),"cm"))

p3 <- plot_grid(p1,p2, ncol = 1, align = "v", rel_heights = c(7,3))

p3

```

### Illustrate cluster 3 split

```{r clust_3_tsne}
# Remove one very distant cell to focus on area of interest on TSNE
cl2 <- colnames(sc_hc@data)[sc_hc@ident==2]
cl2 <- cl2[cl2!="AGGGATGCATTTCACT-1"]

FeaturePlot(sc_hc, c("Actg2","Des","Kit","Ano1"), 
            cells.use = cl2, pt.size = 1, cols.use = c("lightgrey","blue"), 
            no.legend = FALSE)

```

## Identify fibroblast subtypes seen in human data

### Subset to fibroblast-like cells only and detect markers

```{r subset_flc_hc}

sc.flc <- SubsetData(sc_hc, cells.use = colnames(sc_hc@data)[sc_hc@ident %in% c(4,5,10:13)])
TSNEPlot(sc.flc, do.label = T)

flc.markers0 <- FindAllMarkers(sc.flc, only.pos = TRUE, min.pct = 0.25, 
                              thresh.use = 0.25, test.use = "roc", return.thresh = 0.3)

```

### Load human dataset and annotate with mouse gene symbols

```{r human_data}

P150409 <- readRDS("data/P150409/hisat2_fc_GRCh38_rel84.analysis.RData")

sc_Hs <- P150409$scData

sc_Hs_mgi <- convertHumanGeneList(featureNames(sc_Hs))

# Where 1-to-many relationship exists between human and mouse IDs
# select mouse gene symbol with highest mean expression
totUMIcounts <- rowSums(as.matrix(sc_hc@data))
sc_Hs_mgi$totUMIs <- sapply(sc_Hs_mgi$MGI.symbol, function(x) totUMIcounts[x])
sc_Hs_mgi <- sc_Hs_mgi[!is.na(sc_Hs_mgi$totUMIs),]
sc_Hs_mgi %>% group_by(Gene.stable.ID) %>% top_n(1,totUMIs) -> sc_Hs_mgi_top
sc_Hs_mgi_top <- sc_Hs_mgi_top[!duplicated(sc_Hs_mgi_top$Gene.stable.ID),]

# Append to human SCESet
fData(sc_Hs)$mgi_symbol <- NA
fData(sc_Hs)[sc_Hs_mgi_top$Gene.stable.ID,"mgi_symbol"] <- sc_Hs_mgi_top$MGI.symbol

```

### Cross-tabulate cluster markers for fibroblast like cells in both datasets

```{r prep_markers}


flc.markers <- flc.markers0[flc.markers0$myAUC > 0.67,]

## Ensure mouse markers are unique and present in the human dataset
flc.markers <- flc.markers[order(flc.markers$myAUC, decreasing = TRUE),]
flc.markers <- flc.markers[!duplicated(flc.markers$gene),]
flc.markers <- flc.markers[flc.markers$gene %in% fData(sc_Hs)$mgi_symbol,]

## Human cluster markers
hs.markers <- P150409$ROCmark
hs.markers$mgi_symbol <- fData(sc_Hs)[rownames(hs.markers),"mgi_symbol"]
hs.markers <- hs.markers[!is.na(hs.markers$mgi_symbol),]
hs.markers <- hs.markers[hs.markers$AUC > 0.65,]
hs.markers <- hs.markers[order(hs.markers$AUC, decreasing = TRUE),]
hs.markers <- hs.markers[!duplicated(hs.markers$mgi_symbol),]

## Data frame for all shared genes showing cluster annotation in both datasets
Ms_Hs <- data.frame(mgi_symbol = sc_Hs_mgi_top$MGI.symbol, 
                    hs_clust = rep("grey",length(sc_Hs_mgi_top$MGI.symbol)),
                    ms_clust = rep(0, length(sc_Hs_mgi_top$MGI.symbol)))

for(i in flc.markers$gene) {
  Ms_Hs$ms_clust[Ms_Hs$mgi_symbol==i] <- as.character(flc.markers$cluster[flc.markers$gene == i])
}

for(i in hs.markers$mgi_symbol) {
  Ms_Hs$hs_clust[Ms_Hs$mgi_symbol==i] <- hs.markers$Group[hs.markers$mgi_symbol == i]
}

Ms_Hs %>% group_by(hs_clust) %>% summarise(n_mark = length(hs_clust)) -> nmark_hs
Ms_Hs %>% group_by(ms_clust) %>% summarise(n_mark = length(ms_clust)) -> nmark_ms

```

```{r marker_crosstab_hc}

# Crosstabulate marker annotation
d_cast <- dcast(Ms_Hs, hs_clust ~ ms_clust, value.var = "ms_clust", fun.aggregate = length)
d_cross <- as.matrix(d_cast[,-1])
row.names(d_cross) <- d_cast$hs_clust
d_cross <- d_cross[c(2,1,3,4,5),order(as.numeric(colnames(d_cross)))]

fisher <- function(i,j) {
  fisher.test(matrix(c(
    d_cross[i,j],
    sum(d_cross[i,]) - d_cross[i,j],
    sum(d_cross[,j]) - d_cross[i,j], 
    sum(d_cross[row.names(d_cross) != i, colnames(d_cross) != j])
    ), nrow = 2, byrow = TRUE))$p.value
}

d_pval <- sapply(colnames(d_cross), function(j) sapply(row.names(d_cross), function(i) fisher(i,j)))

# Remove unassigned groups from result
d_cross <- d_cross[row.names(d_cross)!="grey", colnames(d_cross)!=0]
d_pval <- d_pval[row.names(d_pval)!="grey", colnames(d_pval)!=0]

# Will display correlations and their p-values
textMatrix = paste(unlist(d_cross), "\n(",
                   signif(unlist(d_pval), 1), ")", sep = "")

dim(textMatrix) = dim(d_cross)

Hs_names <- list(brown = "MF",blue = "Str2", turquoise = "Str3", yellow = "Str1")

par(mar = c(8, 8, 4, 3), mgp = c(5,1,0));
# Display the correlation values within a heatmap plot
labeledHeatmap(Matrix = -log(d_pval),
               xLabels = sapply(colnames(d_cross), function(x) paste0("Cls ",x," (",nmark_ms$n_mark[nmark_ms$ms_clust==x],")")),
               yLabels = paste0("XX",row.names(d_cross)),
               ySymbols = sapply(row.names(d_cross), function(x) paste0(Hs_names[[x]]," (",nmark_hs$n_mark[nmark_hs$hs_clust==x],")")),
               xSymbols = colnames(d_cross),
               xLabelsPosition = "bottom",
               xLabelsAngle = 45,
               xLabelsAdj = 1,
               colors = blueWhiteRed(50)[25:50],
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 1,
               cex.lab = 1,
               zlim = c(0,50),
               plotLegend = TRUE,
               xlab = "Mouse",
               ylab = "Human",
               main = paste0("Marker gene cross-tabulation\n", nrow(Ms_Hs), " shared genes"))

```

### Subset marker genes common to mouse and human

```{r common_markers, fig.width=8, fig.height=4}

Ms_Hs$both <- paste0(Ms_Hs$hs_clust,Ms_Hs$ms_clust)
flc.markers[flc.markers$gene %in% Ms_Hs[Ms_Hs$both=="brown4","mgi_symbol"],]
flc.markers[flc.markers$gene %in% Ms_Hs[Ms_Hs$both=="blue5","mgi_symbol"],]
flc.markers[flc.markers$gene %in% Ms_Hs[Ms_Hs$both=="turquoise10","mgi_symbol"],]
flc.markers[flc.markers$gene %in% Ms_Hs[Ms_Hs$both=="turquoise11","mgi_symbol"],]
flc.markers[flc.markers$gene %in% Ms_Hs[Ms_Hs$both=="yellow12","mgi_symbol"],]
flc.markers[flc.markers$gene %in% Ms_Hs[Ms_Hs$both=="yellow13","mgi_symbol"],]

FeaturePlot(sc_hc,c("Aoc3","Acta2"), cols.use = c("lightgrey","blue"), pt.size = 0.6)
FeaturePlot(sc_hc,c("Sox6","Wnt5a"), cols.use = c("lightgrey","blue"), pt.size = 0.6)
FeaturePlot(sc_hc,c("Maff","Cd55"), cols.use = c("lightgrey","blue"), pt.size = 0.6)
FeaturePlot(sc_hc,c("Ier3","Ptgs2"), cols.use = c("lightgrey","blue"), pt.size = 0.6)
FeaturePlot(sc_hc,c("Cxcl12","Adamdec1"), cols.use = c("lightgrey","blue"), pt.size = 0.6)
FeaturePlot(sc_hc,c("Edil3","Hapln1"), cols.use = c("lightgrey","blue"), pt.size = 0.6)


FeaturePlot(sc_hc,c("Zeb1","Zeb2"), cols.use = c("lightgrey","blue"), pt.size = 0.6)
FeaturePlot(sc_hc,c("Bmp2","Frzb"), cols.use = c("lightgrey","blue"), pt.size = 0.6)
FeaturePlot(sc_hc,c("Thy1","F3"), cols.use = c("lightgrey","blue"), pt.size = 0.6)

```

## Annotated clustering hierarchy

```{r anno_healthy_tsne}

sc_hc <- StashIdent(sc_hc, save.name = "merged_ids")

current.cluster.ids <- 1:13
new.cluster.ids <- c("Epithelial",
                     "Sm mus/ICC",
                     "Pericyte",
                     "Myofibroblast",
                     "Str2",
                     "Endothelial",
                     "Lymphatic",
                     "Glial",
                     "Immune",
                     "Str3",
                     "Str1.3",
                     "Str1.1",
                     "Str1.2")

sc_hc@ident <- plyr::mapvalues(sc_hc@ident, from = current.cluster.ids, to = new.cluster.ids)
flc.markers0$cluster <- plyr::mapvalues(flc.markers0$cluster, from = current.cluster.ids, to = new.cluster.ids)

TSNEPlot(sc_hc, do.label = T, pt.size = 0.6, do.return = TRUE) + xlim(c(-40,50))

sc_hc <- StashIdent(sc_hc, save.name = "clust_names")

```

```{r labelled_tree, fig.height=5}

sc_hc <- BuildClusterTree(sc_hc, do.reorder = F, reorder.numeric = F, do.plot = F)
PlotClusterTree(sc_hc)

```

# DSS colitis dataset

## Data import and setup

```{r load_dss}

# Load the preprocessed seurat object
sc_dss <- readRDS(paste0("output/clust_analysis/", params$data2, "/sc_reclustered.RData"))
clan_dss <- readRDS(paste0("output/clust_analysis/", params$data2, ".RData"))

```

## Identify and merge low confidence nodes

```{r node_confidence_dss, fig.width=6, fig.height=4}

PlotClusterTree(sc_dss)

ns_df <- clan_dss$node.scores
ns_df$node <- factor(ns_df$node, levels = ns_df$node)
ggplot(data=ns_df, aes(x=node, y=oobe)) + geom_col() + geom_hline(yintercept = 0.05, linetype = "dashed")

```

```{r TSNE_split_dss, fig.width=6, fig.height=4}


ColorTSNESplit(sc_dss, node = 17)
ColorTSNESplit(sc_dss, node = 23)

```

```{r merge_low_conf_dss}

sc_dss <- MergeNode(sc_dss, node.use = 23)
sc_dss <- BuildClusterTree(sc_dss, do.reorder = T, reorder.numeric = T, do.plot = F)
TSNEPlot(sc_dss, do.label = T, pt.size = 0.6)

```

## Use healthy cluster definitions to classify cells in DSS

```{r forest_classify, fig.width=8, fig.height=4.5}

hc.rf <- BuildRFClassifier(sc_hc, training.genes = sc_hc@var.genes, training.classes = sc_hc@data.info$merged_ids)

dss_hc_id <- ClassifyCells(sc_hc, hc.rf, new.data = sc_dss@data)
sc_dss <- StashIdent(sc_dss, save.name = "merged_ids")
sc_dss <- SetIdent(sc_dss, ident.use = dss_hc_id)
sc_dss <- StashIdent(sc_dss, save.name = "dss_hc_id")
sc_dss@ident <- plyr::mapvalues(sc_dss@ident, from = current.cluster.ids, to = new.cluster.ids)

p5 <- TSNEPlot(sc_hc, do.label = F, do.return = T, pt.size = 0.4, no.legend = TRUE) + ggtitle("Healthy") + theme(axis.title = element_blank())
p6 <- TSNEPlot(sc_dss, do.label = F, do.return = T, pt.size = 0.4, no.legend = FALSE) + ggtitle("DSS") + theme(axis.title = element_blank())
plot_grid(p5,p6,align = "h", ncol = 2, rel_widths = c(4.15,5.85))

```

### Changes in relative abundance of stromal subtypes

```{r dss_abudance}
cell_df <- data.frame(prop.table(table(sc_hc@data.info$merged_ids)), prop.table(table(dss_hc_id)))
cell_df <- cell_df[,c(1,2,4)]
colnames(cell_df) <- c("cell_type","healthy","dss")
cell_df <- cell_df[order(as.numeric(as.character(cell_df$cell_type))),]
cell_df$cell_type <- plyr::mapvalues(cell_df$cell_type, from = current.cluster.ids, to = new.cluster.ids)
cell_df$fill <- rainbow_hcl(13)
ordered_cts <- cell_df$cell_type[order(cell_df$healthy)]
cell_df$cell_type <- factor(cell_df$cell_type, levels = ordered_cts)
cell_df <- reshape2::melt(cell_df, variable.name = "state", value.name = "proportion")

ggplot(data = cell_df, aes(x = cell_type, label = cell_type, fill = fill, y = proportion))+
  geom_bar(stat = "identity", colour = "black", alpha = 1) + 
  geom_text(aes(label=sprintf("%1.0f%%", 100*proportion)),hjust=-0.2, size = 6) +
  facet_grid(. ~ state) + 
  scale_fill_identity() +
  expand_limits(y = c(0,max(0.5,max(cell_df$proportion) + 0.1))) +
  coord_flip() +
  theme_classic() + 
  theme(
    strip.text.x = element_text(size = 16),
    strip.background = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA),
    axis.ticks.x=element_blank(),
    axis.ticks.y=element_blank(),
    axis.line.x=element_blank(), 
    axis.line.y=element_blank(),
    axis.title.x=element_blank(), 
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.text.y=element_text(size = 16)
  )

```


```{r abundance_byrep}

pr_list <- c(
  lapply(levels(sc_hc@data.info$batch),
         function(x) prop.table(table(sc_hc@data.info$merged_ids[sc_hc@data.info$batch == x]))),
  lapply(levels(sc_dss@data.info$batch),
         function(x) prop.table(table(dss_hc_id[sc_dss@data.info$batch == x])))
  )

pr_mat <- sapply(1:length(pr_list), function(j) {
  sapply(levels(as.factor(sc_hc@data.info$merged_ids)), function(i) {
    pr_list[[j]][i]
  })
})

rownames(pr_mat) <- levels(as.factor(sc_hc@data.info$merged_ids))
colnames(pr_mat) <- c(levels(sc_hc@data.info$batch),levels(sc_dss@data.info$batch))
pr_mat[is.na(pr_mat)] <- 0

cell_df <- data.frame(pr_mat)
cell_df$cell_type <- plyr::mapvalues(rownames(cell_df), from = current.cluster.ids, to = new.cluster.ids)
cell_df <- cell_df[order(as.numeric(rownames(cell_df))),]
cell_df$fill <- rainbow_hcl(13)
ordered_cts <- cell_df$cell_type[order(cell_df$HC1)]
cell_df$cell_type <- factor(cell_df$cell_type, levels = ordered_cts)

cell_tt <- apply(pr_mat,1,function(x) t.test(x[1:3],x[4:6],alternative = "two.sided",paired = FALSE)$p.value)
print(paste(plyr::mapvalues(names(cell_tt), 
                            from = current.cluster.ids, 
                            to = new.cluster.ids),signif(cell_tt,2)))

cell_df <- reshape2::melt(cell_df, variable.name = "state", value.name = "proportion")
cell_df$state <- factor(substring(as.character(cell_df$state),1,2), levels=c("HC","DS"))
levels(cell_df$state) <- c("HC","DSS")

ggplot(data = cell_df[as.numeric(cell_df$cell_type) <= 7,],
       aes(x = state, label = cell_type, fill = fill, y = proportion)) +
  geom_bar(stat = "summary", fun.y = "mean", colour = "black", alpha = 1) + 
  geom_point() +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
        geom="errorbar", width=0.5) +
  facet_grid(. ~ cell_type) + 
  scale_fill_identity() +
  theme_classic() + 
  theme(
    strip.text.x = element_text(size = 12, angle = 0),
    strip.background = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA),
    axis.ticks.x=element_blank(),
    axis.ticks.y=element_blank(),
    axis.line.x=element_blank(), 
    axis.line.y=element_blank(),
    axis.title.x=element_blank(), 
    axis.title.y=element_blank(),
    axis.text.x=element_text(size = 12),
    axis.text.y=element_text(size = 12)
  )

ggplot(data = cell_df[as.numeric(cell_df$cell_type) > 7,], 
       aes(x = state, label = cell_type, fill = fill, y = proportion)) +
  geom_bar(stat = "summary", fun.y = "mean", colour = "black", alpha = 1) + 
  geom_point() +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
        geom="errorbar", width=0.5) +
  facet_grid(. ~ cell_type) + 
  scale_fill_identity() +
  theme_classic() + 
  theme(
    strip.text.x = element_text(size = 12, angle = 0),
    strip.background = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA),
    axis.ticks.x=element_blank(),
    axis.ticks.y=element_blank(),
    axis.line.x=element_blank(), 
    axis.line.y=element_blank(),
    axis.title.x=element_blank(), 
    axis.title.y=element_blank(),
    axis.text.x=element_text(size = 12),
    axis.text.y=element_text(size = 12)
  )

```

## Human UC stromal 4 cluster vs Murine DSS dataset

### Subset to fibroblast-like cells and detect markers

```{r subset_flc_dss}
# Switch back to numeric identifiers
sc_dss <- SetIdent(sc_dss, ident.use = sc_dss@data.info$merged_ids)

dss.flc <- SubsetData(sc_dss, cells.use = colnames(sc_dss@data)[sc_dss@data.info$merged_ids %in% 4:11])
TSNEPlot(dss.flc, do.label = T)

dss.markers0 <- FindAllMarkers(dss.flc, only.pos = TRUE, min.pct = 0.25, thresh.use = 0.25, test.use = "roc", return.thresh = 0.3)

```

```{r human_uc_data}

P160099 <- readRDS("data/P160099/hisat2_fc_GRCh38_rel84.analysis.RData")

sc_UC <- P160099$scData
sc_UC_mgi <- convertHumanGeneList(featureNames(sc_UC))

# Where 1-to-many relationship exists between human and mouse IDs
# select mouse gene symbol with highest mean expression
totUMIcounts <- rowSums(as.matrix(sc_dss@data))
sc_UC_mgi$totUMIs <- sapply(sc_UC_mgi$MGI.symbol, function(x) totUMIcounts[x])
sc_UC_mgi <- sc_UC_mgi[!is.na(sc_UC_mgi$totUMIs),]
sc_UC_mgi %>% group_by(Gene.stable.ID) %>% top_n(1,totUMIs) -> sc_UC_mgi_top
sc_UC_mgi_top <- sc_UC_mgi_top[!duplicated(sc_UC_mgi_top$Gene.stable.ID),]

# Append to human SCESet
fData(sc_UC)$mgi_symbol <- NA
fData(sc_UC)[sc_UC_mgi_top$Gene.stable.ID,"mgi_symbol"] <- sc_UC_mgi_top$MGI.symbol

```

### Cross-tabulate cluster markers

```{r prep_markers_DSS}

dss.markers <- dss.markers0[dss.markers0$myAUC > 0.67,]

## Ensure mouse markers are unique and present in the human dataset
dss.markers <- dss.markers[order(dss.markers$myAUC, decreasing = TRUE),]
dss.markers <- dss.markers[!duplicated(dss.markers$gene),]
dss.markers <- dss.markers[dss.markers$gene %in% fData(sc_UC)$mgi_symbol,]

## Human cluster markers
uc.markers <- P160099$ROCmark
uc.markers$mgi_symbol <- fData(sc_UC)[rownames(uc.markers),"mgi_symbol"]
uc.markers <- uc.markers[!is.na(uc.markers$mgi_symbol),]
uc.markers <- uc.markers[uc.markers$AUC > 0.65,]
uc.markers <- uc.markers[order(uc.markers$AUC, decreasing = TRUE),]
uc.markers <- uc.markers[!duplicated(uc.markers$mgi_symbol),]

## Data frame for all shared genes showing cluster annotation in both datasets
dss_uc <- data.frame(mgi_symbol = sc_UC_mgi_top$MGI.symbol, 
                    hs_clust = rep("grey",length(sc_UC_mgi_top$MGI.symbol)),
                    ms_clust = rep(0, length(sc_UC_mgi_top$MGI.symbol)))

for(i in dss.markers$gene) {
  dss_uc$ms_clust[dss_uc$mgi_symbol==i] <- as.character(dss.markers$cluster[dss.markers$gene == i])
}

for(i in uc.markers$mgi_symbol) {
  dss_uc$hs_clust[dss_uc$mgi_symbol==i] <- uc.markers$Group[uc.markers$mgi_symbol == i]
}

dss_uc %>% group_by(hs_clust) %>% summarise(n_mark = length(hs_clust)) -> nmark_uc
dss_uc %>% group_by(ms_clust) %>% summarise(n_mark = length(ms_clust)) -> nmark_dss

```

```{r marker_crosstab_dss, fig.height=3}

# Crosstabulate marker annotation
d_cast <- dcast(dss_uc, hs_clust ~ ms_clust, value.var = "ms_clust", fun.aggregate = length)
d_cross <- as.matrix(d_cast[,-1])
row.names(d_cross) <- d_cast$hs_clust
d_cross <- d_cross[c(3,2,6,1,5,4),order(as.numeric(colnames(d_cross)))]

fisher <- function(i,j) {
  fisher.test(matrix(c(
    d_cross[i,j],
    sum(d_cross[i,]) - d_cross[i,j],
    sum(d_cross[,j]) - d_cross[i,j], 
    sum(d_cross[row.names(d_cross) != i, colnames(d_cross) != j])
    ), nrow = 2, byrow = TRUE))$p.value
}

d_pval <- sapply(colnames(d_cross), function(j) sapply(row.names(d_cross), function(i) fisher(i,j)))

# Remove grey module from result
d_cross <- d_cross[row.names(d_cross)!="grey", colnames(d_cross)!=0]
d_pval <- d_pval[row.names(d_pval)!="grey", colnames(d_pval)!=0]

# Will display correlations and their p-values
textMatrix = paste(unlist(d_cross), "\n(",
                   signif(unlist(d_pval), 1), ")", sep = "")

dim(textMatrix) = dim(d_cross)

UC_names <- list(brown = "MF",blue = "Str2", turquoise = "Str3", yellow = "Str1", green = "Str4")

par(mar = c(6, 12, 4, 3), mgp = c(3,1,0));
# Display the correlation values within a heatmap plot
labeledHeatmap(Matrix = -log(d_pval[1,,drop=FALSE]),
               xLabels = sapply(colnames(d_cross), function(x) paste0("Cls ",x,"\n(",nmark_dss$n_mark[nmark_dss$ms_clust==x],")")),
               yLabels = paste0("XX",row.names(d_cross))[1],
               ySymbols = sapply(row.names(d_cross), function(x) paste0(UC_names[[x]]," (",nmark_uc$n_mark[nmark_uc$hs_clust==x],")"))[1],
               xSymbols = colnames(d_cross),
               xLabelsPosition = "bottom",
               xLabelsAngle = 0,
               xLabelsAdj = c(0.5,1),
               colors = blueWhiteRed(50)[25:50],
               textMatrix = textMatrix[1,,drop=FALSE],
               setStdMargins = FALSE,
               cex.text = 1,
               cex.lab = 1,
               zlim = c(0,20),
               plotLegend = TRUE,
               xlab = "Mouse",
               ylab = "Human",
               main = paste0("Marker gene cross-tabulation\n", nrow(dss_uc), " shared genes"))

```

```{r cl7_split, fig.height=4, fig.width=4}

sc_dss@data.info$Str4 <- sc_dss@data.info$merged_ids == 7
TSNEPlot(sc_dss, group.by = "Str4",colors.use = c("lightgrey","red"), no.legend = TRUE, do.return = TRUE) + 
  theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank())

```

```{r s4_subset, fig.width=8, fig.height=4}

dss_uc$both <- paste0(dss_uc$hs_clust,dss_uc$ms_clust)
dss.markers[dss.markers$gene %in% dss_uc[dss_uc$both=="green7","mgi_symbol"],]

FeaturePlot(dss.flc,c("Il33","Ccl19"), cols.use = c("lightgrey","blue"), pt.size = 0.6)

```

```{r clust_heat}

pr_tab <- sapply(levels(as.factor(sc_dss@data.info$merged_ids)), function(x) prop.table(table(dss_hc_id[sc_dss@data.info$merged_ids==x])))
class(pr_tab)
rownames(pr_tab) <- plyr::mapvalues(rownames(pr_tab), from = current.cluster.ids, to = new.cluster.ids)
pr_tab <- pr_tab[as.hclust(sc_hc@cluster.tree[[1]])$labels,as.hclust(sc_dss@cluster.tree[[1]])$labels]
heatmap.2(pr_tab, Colv = as.dendrogram(as.hclust(sc_dss@cluster.tree[[1]])), 
          Rowv = as.dendrogram(as.hclust(sc_hc@cluster.tree[[1]])), 
          trace = "none", col = viridis(75),xlab = "DSS cluster", ylab = "Healthy cluster", 
          density.info = "none", key.title = NA, margins = c(4,8))

```

```{r dss_labelled}

sc_dss <- SetIdent(sc_dss, ident.use = sc_dss@data.info$merged_ids)
current.cluster.ids <- 1:11
new.cluster.ids <- c("Lymphatic",
                     "Endothelial",
                     "Pericyte",
                     "Str3.1",
                     "Str3.2",
                     "Str3.3",
                     "Str4",
                     "Myofibroblast",
                     "Str1.1",
                     "Str2",
                     "Str1.2")

sc_dss@ident <- plyr::mapvalues(sc_dss@ident, from = current.cluster.ids, to = new.cluster.ids)
dss.markers0$cluster <- plyr::mapvalues(dss.markers0$cluster, from = current.cluster.ids, to = new.cluster.ids)

TSNEPlot(sc_dss, do.label = T, pt.size = 0.6, do.return = TRUE)

sc_dss <- StashIdent(sc_dss, save.name = "clust_names")

```

```{r labelled_tree_dss, fig.height=5}

sc_dss <- BuildClusterTree(sc_dss, do.reorder = F, reorder.numeric = F, do.plot = F)
PlotClusterTree(sc_dss)

```

## Cell cycle stage comparison

```{r cell_cycle_clust}

sc_dss@ident <- plyr::mapvalues(sc_dss@ident, from = c("Str3.1","Str3.2","Str3.3","Str1.1","Str1.2","Str1.3"), 
                                to = c(rep("Str3",3),rep("Str1",3)), warn_missing = FALSE)
sc_hc@ident <- plyr::mapvalues(sc_hc@ident, from = c("Str1.1","Str1.2","Str1.3"), to = rep("Str1",3), warn_missing = FALSE)

sc_dss <- StashIdent(sc_dss, save.name = "no_subclust")
sc_hc <- StashIdent(sc_hc, save.name = "no_subclust")

cc_df <- data.frame(cluster = sc_dss@ident, G2M = sc_dss@data.info$G2M > 0.5 & sc_dss@data.info$G2M > sc_dss@data.info$G1)
cc_df %>% group_by(cluster) %>% summarise(prop = sum(G2M)/length(G2M)) -> cc_dss

cc_df <- data.frame(cluster = sc_hc@ident, G2M = sc_hc@data.info$G2M > 0.5 & sc_hc@data.info$G2M > sc_hc@data.info$G1)
cc_df %>% group_by(cluster) %>% summarise(prop = sum(G2M)/length(G2M)) -> cc_hc

cc_dss <- as.data.frame(cc_dss)
cc_hc <- as.data.frame(cc_hc)
cc_all <- data.frame(cluster = c("Endothelial", "Lymphatic","Pericyte","Epithelial",
                                 "Sm mus/ICC","Glial","Immune","Myofibroblast",
                                 "Str1","Str2","Str3","Str4"), Healthy = NA, DSS = NA)

cc_all$Healthy <- sapply(cc_all$cluster, function(x) ifelse(x %in% cc_hc$cluster, 100*cc_hc$prop[cc_hc$cluster==x], NA))
cc_all$DSS <- sapply(cc_all$cluster, function(x) ifelse(x %in% cc_dss$cluster, 100*cc_dss$prop[cc_dss$cluster==x], NA))
cc_all$cluster <- factor(cc_all$cluster, levels = cc_all$cluster)
cc_all <- melt(cc_all,variable.name = "condition",value.name = "pct_G2M")
cc_all$isNA <- ifelse(is.na(cc_all$pct_G2M),"nd","")
cc_all$pct_G2M[is.na(cc_all$pct_G2M)] <- 0

ggplot(data=cc_all, aes(x=cluster, y=pct_G2M, fill=cluster)) + geom_col() + geom_text(aes(label = isNA, y=0), col = "darkgrey", vjust = 0) + 
  facet_grid(condition ~ ., scales = "free_x") + 
  scale_fill_discrete(guide = "none") +
  ylab("% cells G2/M phase") +
  ggtitle("Proliferative activity by cluster") +
  theme_bw() +
  theme(axis.title = element_text(size = 16),
        axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
        strip.text.y = element_text(size = 16))
```

```{r cell_cycle_cond}

ggplot(data=sc_hc@data.info[,c("G1","G2M","phases")], aes(x=G1,y=G2M,fill=phases)) + 
  geom_point(pch = 21, size = 4, alpha = 0.7) +
  xlim(c(0,1)) +
  ylim(c(0,1)) +
  geom_segment(aes(x = 0.5, y = 0, xend = 0.5, yend = 0.5)) +
  geom_segment(aes(x = 0, y = 0.5, xend = 0.5, yend = 0.5)) +
  geom_segment(aes(x = 0.5, y = 0.5, xend = 1, yend = 1)) +
  annotate("text", x = 0.4, y = 0.4, label = "S") +
  annotate("text", x = 0.5, y = 0.6, label = "G2/M") +
  annotate("text", x = 0.6, y = 0.5, label = "G0/G1") +
  ggtitle("Healthy controls")

ggplot(data=sc_dss@data.info[,c("G1","G2M","phases")], aes(x=G1,y=G2M,fill=phases)) + 
  geom_point(pch = 21, size = 4, alpha = 0.7) +
  xlim(c(0,1)) +
  ylim(c(0,1)) +
  geom_segment(aes(x = 0.5, y = 0, xend = 0.5, yend = 0.5)) +
  geom_segment(aes(x = 0, y = 0.5, xend = 0.5, yend = 0.5)) +
  geom_segment(aes(x = 0.5, y = 0.5, xend = 1, yend = 1)) +
  annotate("text", x = 0.4, y = 0.4, label = "S") +
  annotate("text", x = 0.5, y = 0.6, label = "G2/M") +
  annotate("text", x = 0.6, y = 0.5, label = "G0/G1") +
  ggtitle("DSS challenge")

```

## Save out objects

```{r save_objects}

save_list <- list()
objects_to_save <- c("dss.markers0","flc.markers0","sc_dss","sc_hc")

for(x in objects_to_save) {
    save_list[[x]] <- get(x)
}

saveRDS(save_list, file = paste0("output/",params$script_name,".RData"))

```

