---
title: '10X normalisation and cluster annotation'
author: "James Kinchen, University of Oxford"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
        code_folding: hide
        dev: [png,pdf]
params:
  script_name:
    input: text
    label: 'Script identifier'
    value: 'norm_clust'
  source_dir:
    input: text
    label: 'Input dataset'
    value: 'data/HC_3L/raw_gene_bc_matrices_mex/mm10_pL63_mCit'
  source_metadata:
    input: text
    label: 'Input metadata from cell detection script'
    value: 'output/celldetection/10X_Ms_Healthy.RData'
  proj_name:
    input: text
    label: 'Seurat dataset name'
    value: '10X_Ms_DSS'
  calc_jackstraw:
    input: text
    label: 'Calculate Jackstraw - takes 1hr'
    value: 'no'
  pcs_use:
    input: numeric
    lable: 'PCs to use for community detection'
    value: 28
  cluster_res:
    input: numeric
    label: 'Clustering resolution'
    value: 1
  tsne_pcs:
    input: numeric
    label: 'PCs for TSNE'
    value: 28
---

```{r setup, include=FALSE, cache=FALSE}

knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 6, message = FALSE, warning = FALSE, cache = TRUE, 
                      cache.path = paste0("cache/",params$script_name,"/",params$proj_name,"/"),
                      fig.path = paste0("output/",params$script_name,"/",params$proj_name,"/"))

```

```{r setup_2, include=FALSE, cache=FALSE}

options(stringsAsFactors = FALSE)
library(Seurat)
library(dplyr)
library(Matrix)
library(colorspace)
library(viridis)
library(stats)
library(scran)

```

# Dataset - `r params$proj_name`

## Data import and setup

### Load unfiltered UMI count matrix

```{r import_raw}

# Load the unfiltered healthy control dataset
raw.data <- Read10X(params$source_dir)

# Load the metadata from the cell detection script
celldetect <- readRDS(params$source_metadata)

```

### Subset cells passing QC

```{r subset_qc}

qcpass.data <- raw.data[,rownames(celldetect$sc_metadata)[!celldetect$sc_metadata$qc_all]]

dim(qcpass.data)

```

### Subset to remove control cells

```{r subset_ctrl}

qcpass.data <- qcpass.data[,!colnames(qcpass.data) %in% celldetect$ctrl_bcs]

dim(qcpass.data)

```

### Construct Seurat object with metadata

```{r build_seurat}
sc_meta <- celldetect$sc_metadata[colnames(qcpass.data),]

# Initialize the Seurat object
sc <- new("seurat", raw.data = qcpass.data)

# Setup Seurat object, apply gene filter (cell filter already done during cell detection)
sc <- Setup(sc, min.cells = 3, min.genes = 200, do.logNormalize = T, 
            total.expr = 1e4, project = params$proj_name, meta.data = sc_meta)

# Dataset dimensions post filter
print(paste(dim(sc@data)[1], "genes"))
print(paste(dim(sc@data)[2], "cellular GEMs"))

```

## Normalisation

### Determine variable genes

```{r variable_genes}

#Identify variable genes
sc <- MeanVarPlot(sc ,fxn.x = expMean, fxn.y = logVarDivMean, x.low.cutoff = 0.0125, x.high.cutoff = 3, y.cutoff = 0.5, do.contour = F)
print(paste(length(sc@var.genes), "variable genes"))

```


### Calculate cell cycle scores

```{r cell_cycle}

# Get pre-trained marker sets
mm.pairs <- readRDS(system.file("exdata", "mouse_cycle_markers.rds", package="scran"))

# Get the expression data matrix
cc_mat <- as.matrix(sc@data)

# Switch gene ids to ensembl to match pre-trained set
ms_ensembl <- read.table(paste0(params$source_dir,"/genes.tsv"),
                         header = FALSE, sep = "\t", quote = "", stringsAsFactors = FALSE)
cc_ensembl <- sapply(rownames(cc_mat), function(x) ms_ensembl$V1[ms_ensembl$V2==x][1])

# Cell cycle anotation using the cyclone pairs method
cyc <- cyclone(cc_mat, mm.pairs, gene.names=cc_ensembl, verbose = TRUE)

# Add back to Seurat object
for(scr in colnames(cyc$scores)) {
  sc@data.info[[scr]] <- cyc$scores[[scr]]
  sc@data.info[[paste0("norm_",scr)]] <- cyc$normalized.scores[[scr]]
}

sc@data.info[["phases"]] <- cyc$phases

```

### Regress out unwanted variation

```{r regress_out}

#Regress out unwanted variation
sc <- RegressOut(sc, latent.vars = c("nUMI","batch","G2M","G1"), 
                        genes.regress = sc@var.genes, model.use = "negbinom")

```

## Dimension reduction

### PCA

```{r PCA}

#Linear dimension reduction
sc <- PCA(sc, pc.genes = sc@var.genes, do.print = FALSE, pcs.store = 50)
sc <- ProjectPCA(sc, do.print = FALSE)

```

### PC evaluation - Scree plot

```{r PC_scree, fig.width = 4, fig.height=3}

p1 <- PCElbowPlot(sc, num.pc = 50)

# Fit line for the last 10 (presumed non-significant) PCs to aid interpretation
bf <- lm(sdev ~ pc, data = p1$data[40:50,])
summary(bf)

p1 + geom_abline(intercept = bf$coefficients[1], slope = bf$coefficients[2], linetype = "dashed") + 
  geom_vline(xintercept = params$pcs_use, linetype = "dotted")

```

### PC evaluation - permutation test

```{r PC_permutation}

if(file.exists(paste0("output/",params$script_name,"/",params$proj_name,"/sc_jackstraw.RData")) & params$calc_jackstraw == "no") {
  sc <- readRDS(file = paste0("output/",params$script_name,"/",params$proj_name,"/sc_jackstraw.RData"))
} else {
  sc <- JackStraw(sc, num.replicate = 100, do.print = TRUE, num.pc = 50)
  saveRDS(sc, file = paste0("output/",params$script_name,"/",params$proj_name,"/sc_jackstraw.RData"))
}

```

```{r PC_permutation_plot, fig.width=4, fig.height=3}

# Sample plots of tested PCs
JackStrawPlot(sc, PCs = paste0("PC",seq(1, 50, by = 6)), plot.y.lim = 1.0, score.thresh = 1e-05)

# Calculate overall p-values for PCs
score.thresh <- 1e-05
pc.score <- sapply(1:50, function(x) {
  prop.test(c(length(which(sc@jackStraw.empP[, x] <= score.thresh)), floor(nrow(sc@jackStraw.empP) * score.thresh)), c(nrow(sc@jackStraw.empP), nrow(sc@jackStraw.empP)))$p.val
})

ggplot(data = data.frame(PC = 1:50, pval = -log10(pc.score)), aes(x=PC, y=pval)) + geom_col() + labs(y="-log10(p value)") +
  geom_vline(xintercept = params$pcs_use, linetype = "dotted")

```

### PC evaluation - inspection

```{r PC_heatmap}

for(i in seq(1, 50, by = 6)) {
  print(PCHeatmap(sc, pc.use = i, cells.use = 100, do.balanced = TRUE, labCol = NA, col.use = viridis(75)))
}

```

### Graph-based clustering

```{r final_clust, cache=FALSE}

print(paste0("PCs used for clustering - 1:",params$pcs_use))
print(paste0("Clustering resolution: ",params$cluster_res))

sc <- FindClusters(sc, pc.use = 1:params$pcs_use, 
                   resolution = params$cluster_res, 
                   print.output = FALSE, save.SNN = TRUE, plot.SNN = FALSE)

```

```{r calc_tsne, cache=FALSE}

print(paste0("PCs used for TSNE - 1:",params$tsne_pcs))

sc <- RunTSNE(sc, dims.use = 1:params$tsne_pcs, do.fast = T)

```

```{r plot_TSNE, cache=FALSE}

TSNEPlot(sc, do.label = T, pt.size = 0.5, do.return = TRUE)

```

### Save out control cell barcodes

```{r save_data, cache = FALSE}
  
saveRDS(sc, file = paste0("output/",params$script_name,"/",params$proj_name,"/sc_clustered.RData"))

```

